---
title: "A5-2"
author: "Swetha Narayanan"
date: "11/18/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up environment
```{r}
library(caret)
library(xgboost)
library(DMwR)
library(ROSE)
library(ROCR)
library(dplyr)
library(tidyverse)

set.seed(5152)

data <- read.csv("data/application_v3.csv")
data$TARGET <- factor(data$TARGET)
levels(data$TARGET) <- c("OTHER", "DIFFICULTY")

#Summary on the target variable - Minority class is 6% of total class labels
summary(data$TARGET)

train_idx <- createDataPartition(data$TARGET, p = 0.7, list = FALSE)
train <- data[train_idx, ]
test  <- data[-train_idx, ]
```

## Set up functions
```{r}
calculate_cm <- function(model, test){
  pred <- predict(model, test)
  cm <- confusionMatrix(pred, test$TARGET, positive = 'DIFFICULTY' )
  return(cm)
}

#We want to calculate AUC for the minority class - i.e target variable in difficulty
calculate_auc <- function(model, test){
  pred_obj <- prediction(predictions = predict(model, test, type='prob')$DIFFICULTY, labels =  test$TARGET)
  perf_auc <- performance(pred_obj, measure = "auc")
  return(perf_auc@y.values)
}

build_tuned_xgb_model <- function(train, ctrl){
  starttime <- Sys.time()
  grid <- expand.grid(.nrounds=c(40,50,60),.eta=c(0.2,0.3,0.4),.gamma=c(0,1),.max_depth=c(2,3,4),.colsample_bytree=c(0.8),.subsample=c(1),.min_child_weight=c(1))
  model <- train(TARGET ~ ., data = train, method = "xgbTree", metric = "ROC", trControl = ctrl, tuneGrid = grid)
  endtime<-Sys.time()
  print(endtime-starttime)
  print(model$bestTune)
  return(model)
}
```

## Baseline model
```{r}
ctrl <- trainControl(method = "cv", number = 10, summaryFunction = twoClassSummary, classProbs = TRUE, allowParallel = TRUE)
model_baseline <- build_tuned_xgb_model(train, ctrl)
#Confusion Matrix
cm_original <- calculate_cm(model_baseline, test)
cm_original
#AUC
auc_original <- calculate_auc(model_baseline, test)
auc_original
```

## Undersampling
### Undersampling randomly downsamples the majority class. 
### Downside: In general, the more imbalanced the dataset the more samples will be discarded when undersampling, therefore ### throwing away potentially useful information
```{r}
ctrl <- trainControl(method = "cv", number = 10, summaryFunction = twoClassSummary, classProbs = TRUE, sampling = "down", allowParallel = TRUE)
model.down <- build_tuned_xgb_model(train, ctrl)
#Confusion Matrix
cm_under <- calculate_cm(model.down, test)
cm_under
#AUC
auc_under <- calculate_auc(model.down, test)
auc_under
```

## Over sampling
### Oversampling randomly replicates minority instances to increase their population. 
### Downside: Replicating data is not without consequence—since it results in duplicate data, it makes variables appear to have lower variance than they do
```{r}
ctrl <- trainControl(method = "cv", number = 10, summaryFunction = twoClassSummary, classProbs = TRUE, sampling = "up", allowParallel = TRUE)
model.up <- build_tuned_xgb_model(train, ctrl)
#Confusion Matrix
cm_over <- calculate_cm(model.up, test)
cm_over
#AUC
auc_over <- calculate_auc(model.up, test)
auc_over
```

## Smote sampling : Synthetic Minority Over-sampling Technique
### SMOTE draws artificial samples by choosing points that lie on the line connecting the rare observation to one of its nearest neighbors in the feature space.
```{r}
ctrl <- trainControl(method = "cv", number = 10, summaryFunction = twoClassSummary, classProbs = TRUE, sampling = "smote", allowParallel = TRUE)
model.smote <- build_tuned_xgb_model(train, ctrl)
#Confusion Matrix
cm_smote <- calculate_cm(model.smote, test)
cm_smote
#AUC
auc_smote <- calculate_auc(model.smote, test)
auc_smote
```

## Under and oversampling
### In this case, the minority class is oversampled with replacement and majority class is undersampled without replacement.
### Unlike SMOTE, we dont create synthetic samples here - we just work with existing data with and without replacement
```{r}
summary(train$TARGET)
summary(test$TARGET)

data_balanced_both <- ovun.sample(TARGET ~ ., data = train, method = "both")
summary(data_balanced_both$data$TARGET)

## Data is in 1:16 : rareclass: commonclass ratio
ctrl <- trainControl(method = "cv", number = 10, summaryFunction = twoClassSummary, classProbs = TRUE, allowParallel = TRUE)
model_under_over <- build_tuned_xgb_model(data_balanced_both$data, ctrl)
#Confusion Matrix
cm_under_over <- calculate_cm(model_under_over, test)
cm_under_over
#AUC
auc_under_over <- calculate_auc(model_under_over, test)
auc_under_over
```

## Comparing performance
### Note that metrics here are calculated for the DIFFICULTY class
```{r}
AUC_metrics <- data.frame(auc_original,auc_under,auc_over,auc_under_over,auc_smote)
colnames(AUC_metrics) <- c("original","under","over", "under_over","smote")
models <- list(original = model_baseline,
               under = model.down,
               over = model.up,
               under_over = model_under_over,
               smote = model.smote
)
resampling <- resamples(models)
bwplot(resampling)

# Nicer plot for comparing results
comparison <- data.frame(model = names(models),
                         Precision = rep(NA, length(models)),
                         Recall = rep(NA, length(models)),
                         F1 = rep(NA, length(models)),
                         AUC = rep(NA, length(models)))

for (name in names(models)) {
  model <- get(paste0("cm_", name))
  metrics <- model$byClass
  
  comparison[comparison$model == name, ] <- filter(comparison, model == name) %>%
    mutate(Precision = metrics["Precision"],
           Recall = metrics["Recall"],
           F1 = metrics["F1"],
           AUC = as.numeric(AUC_metrics[name]))
}

comparison %>%
  gather(x, y, Precision:AUC) %>%
  ggplot(aes(x = x, y = y, color = model)) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 3)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
