---
title: "Assignment3"
author: "Swetha Narayanan"
date: "10/7/2018"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```

## Problem Statement - Understanding how ensemble methods improve performance
The data is a simulated dataset with one binary label and 15 numerical features. There are 2000 records in the training data (with label values) and 2000 records in the test data (without label values).

```{r}
set.seed(42)

A3_train <- read.csv("A3_train.csv")
A3_test <- read.csv("A3_test.csv")
A3_test_y <- read.csv("A3_test_y_true.csv")


#split into train and validation
train_data <- A3_train[1:1500,]
validation_data <- A3_train[1501:2000,]
test_data <- A3_test

train_data$y <- as.factor(train_data$y)
validation_data$y <- as.factor(validation_data$y)
```

## Task 1 - Random Forest of a post-pruned rpart
Performance metric is simple accuracy for binary labels.
```{r}
set.seed(42)
library(rpart)

# Train random forest model manually
train_random_forest <- function(n_trees, n_features, training_data, target_col_name) {
  models <- lapply(1:n_trees, function(i) {
    
    # bootstrapping
    n_samples <- nrow(training_data)
    sample_row_ids <- sample(1:n_samples,replace=TRUE)
    new_training_data <- training_data[sample_row_ids, ]
    
    #column sampling
    target_var_index <- which(colnames(new_training_data)==target_col_name)
    feature_indices <- c(1:ncol(new_training_data))[-target_var_index]
    sample_cols <- sample(feature_indices, n_features)
    new_training_data <- new_training_data[append(sample_cols,target_var_index)]
    
    #build model
    formula <- as.formula(paste(target_col_name, '~.'))
    new_model <- rpart(formula, data=new_training_data)
    
    #Prune model
    bestcp <- new_model$cptable[which.min(new_model$cptable[,"xerror"]),"CP"]
    pruned_model <- prune(new_model, cp = bestcp)
    return(pruned_model)
  })
  return(models)
}

predict_random_forest <- function(models, test_data) {
  preds <- sapply(models, function(model) {
    return(predict(model, test_data, type = 'class'))
  })
   # apply majority voting
  voted_pred <- apply(preds, 1, function(row) {
    return(names(which.max(table(row))))
  })
  return(as.numeric(voted_pred))
}

#Typically, for a classification problem with p features, √p (rounded down) features are used in each split.
#Source : https://en.wikipedia.org/wiki/Random_forest#cite_note-elemstatlearn-3
models_rf <- train_random_forest(50, sqrt(ncol(train_data)), train_data, 'y')
train_pred_rf <- predict_random_forest(models_rf, train_data)
mean(train_pred_rf == as.numeric(train_data$y)-1) 
validation_pred_rf <- predict_random_forest(models_rf, validation_data)
mean(validation_pred_rf == as.numeric(validation_data$y)-1) 
```
## Task 2 - Manual Stacking
Stacking of three algorithms: C50 with default parameter values, KNN with k=3, and your random forest in Task 1. The output of level0 is a binary label (not predicted probability). Logistic regression is used for the level1 algorithm. The final output is a binary label and the performance metric is simple accuracy.
```{r}
set.seed(42)
#split into train and validation and test
train_l0_data <- A3_train[1:1000,]
train_l1_data <- A3_train[1001:1700,]
test_data <- A3_train[1701:2000,]

train_l0_data$y <- as.factor(train_l0_data$y)
train_l1_data$y <- as.factor(train_l1_data$y)
test_data$y <- as.factor(test_data$y)

#Model 1
library(C50)
model_tree <- C5.0(y ~ ., train_l0_data)
tree_accuracy_l0 <- mean(predict(model_tree, train_l0_data) == train_l0_data$y)

#Model 2
library(class)
train <- train_l0_data[-which(colnames(train_l0_data)=='y')]
pred_knn <- knn(train, train, train_l0_data$y, k = 3)
knn_accuracy_l0 <- mean(pred_knn == train_l0_data$y)

#Model 3: from Q1
model_rf <- train_random_forest(50, sqrt(ncol(train_l0_data)), train_l0_data, 'y')
rf_accuracy_l0 <- mean(predict_random_forest(model_rf, train_l0_data) == train_l0_data$y)

#Prediction on l1 training data
pred_tree <- predict(model_tree, train_l1_data)
tree_accuracy_l1 <- mean(pred_tree == train_l1_data$y)  #Calculate Tree accuracy

pred_knn <- knn(train, train_l1_data[-which(colnames(train_l1_data)=='y')], train_l0_data$y, k = 3)
knn_accuracy_l1 <- mean(pred_knn == train_l1_data$y) #Calculate KNN accuracy

pred_rf <- predict_random_forest(model_rf, train_l1_data)
rf_accuracy_l1 <- mean(pred_rf == train_l1_data$y) #Calculate RF accuracy

#Generic meta classifier 
build_meta_classifier <- function( stacked_data, outcome_variable){
  stacked_model <- glm(outcome_variable ~ ., stacked_data, family = "binomial")
  return(stacked_model)
} 

stacked_data <- data.frame("tree" = pred_tree, "knn" = pred_knn, "rf"= pred_rf, outcome_variable = train_l1_data$y, row.names = NULL)
stacked_model <- build_meta_classifier( stacked_data, "y" )

#Stacking training accuracy
prob_stacked <- predict(stacked_model, stacked_data, type = "response")
pred_stacked <- as.factor(ifelse(prob_stacked >= 0.5, 1, 0))
stacking_accuracy_l1 <- mean(pred_stacked == train_l1_data$y)

c(rf_accuracy_l1, knn_accuracy_l1, tree_accuracy_l1, stacking_accuracy_l1)

#Stacking test accuracy
pred_tree <- predict(model_tree, test_data) 
tree_accuracy_test <- mean(pred_tree == test_data$y)  #Calculate Tree accuracy

pred_knn <- knn(train, test_data[-which(colnames(test_data)=='y')], train_l0_data$y, k = 3) 
knn_accuracy_test <- mean(pred_knn == test_data$y) #Calculate KNN accuracy

pred_rf <- predict_random_forest(models_rf, test_data)
rf_accuracy_test <- mean(pred_rf == test_data$y) #Calculate RF accuracy

stacked_data <- data.frame("tree" = pred_tree, "knn" = pred_knn, "rf"= pred_rf )
prob_stacked <- predict(stacked_model, stacked_data, type = "response")
pred_stacked <- as.factor(ifelse(prob_stacked >= 0.5, 1, 0))
stacking_accuracy_test <- mean(pred_stacked == test_data$y)

c(rf_accuracy_test, knn_accuracy_test, tree_accuracy_test, stacking_accuracy_test)
```

## Task 3 - Data competition - stacking usng ROC as metric
```{r}
set.seed(42)
library("caretEnsemble")
library("caret")
library("ROCR")
library("corrplot")
library("MASS")

A3_train <- read.csv("A3_train.csv")
A3_test <- read.csv("A3_test.csv")

#Data visualization, x1, x5 , x8, x14 seem to be highly correlated. So we can drop 3 of them
corrplot(cor(A3_train), method = "circle")
A3_train <- A3_train [-c(1,5,14)]
A3_test <- A3_test[-c(1,5,14)]

#Feature Engineering - x4*x8 seems to be signficant through stepAIC, so we add it to the model
glm1 <- glm(y ~ .+ x4 *x8, data=A3_train, family=binomial())
glm.best <- stepAIC(glm1, direction="both",trace=0)
glm.best
A3_train$x16 = A3_train$x4 * A3_train$x8
A3_test$x16 = A3_test$x4 * A3_test$x8

#split into train and validation
train_data <- A3_train[1:1500,]
validation_data <- A3_train[1501:2000,]
test_data <- A3_test

train_data$y <- as.factor(train_data$y)
levels(train_data$y) <- make.names(levels(train_data$y))
validation_data$y <- as.factor(validation_data$y)
levels(validation_data$y) <- make.names(levels(validation_data$y))

#stratified sampling 
folds <- createFolds(train_data$y , k = 5)
#We specify summaryFunction so that we can use ROC as our metric for primary model selection later
control <- trainControl(method='repeatedcv', number=5, repeats=3, index=folds, savePredictions='final', classProbs=TRUE, summaryFunction=twoClassSummary)
algos <- c('rf', 'xgbTree', 'C5.0', 'nnet', 'glmStepAIC')
models <- caretList(y ~ . , data=train_data, trControl=control, metric='ROC', methodList=algos)
models_perf <- resamples(models)
summary(models_perf)
modelCor(models_perf)

#Now, let's use xgboost as our meta model
stack_control <- trainControl(method='repeatedcv', number=5, index=folds, repeats=3, classProbs=TRUE)
stack_xgb <- caretStack(models, method='xgbTree', metric='ROC', trControl=stack_control,trace = FALSE )
stack_xgb

#Predict on training data
stacking_prob_train <- 1 - predict(stack_xgb, train_data, type = "prob") #Gives probability of y = 1
stacking_pred_train <- prediction(stacking_prob_train, labels=as.numeric(train_data$y)-1)
performance(stacking_pred_train, 'auc')@y.values #T rainingAUC
mean(predict(stack_xgb, train_data, type = "raw") == train_data$y) #Training Accuracy

#Predict on validation data
rf_prob_validation <- predict(models$rf, validation_data, type = "prob")[,"X1"] #Gives predictedprobability of y = 1
rf_pred_validation <- prediction(rf_prob_validation, labels=as.numeric(validation_data$y)-1)
performance(rf_pred_validation, 'auc')@y.values #RF Validation AUC

xbg_prob_validation <- predict(models$xgbTree, validation_data, type = "prob")[,"X1"] #Gives predictedprobability of y = 1
xgb_pred_validation <- prediction(xbg_prob_validation, labels=as.numeric(validation_data$y)-1)
performance(xgb_pred_validation, 'auc')@y.values #XGB Validation AUC

c50_prob_validation <- predict(models$C5.0, validation_data, type = "prob")[,"X1"] #Gives predictedprobability of y = 1
c50_pred_validation <- prediction(c50_prob_validation, labels=as.numeric(validation_data$y)-1)
performance(c50_pred_validation, 'auc')@y.values #C50 Validation AUC

nnet_prob_validation <- predict(models$nnet, validation_data, type = "prob")[,"X1"] #Gives predictedprobability of y = 1
nnet_pred_validation <- prediction(nnet_prob_validation, labels=as.numeric(validation_data$y)-1)
performance(nnet_pred_validation, 'auc')@y.values #NNet Validation AUC

glm_prob_validation <- predict(models$glmStepAIC, validation_data, type = "prob")[,"X1"] #Gives predictedprobability of y = 1
glm_pred_validation <- prediction(glm_prob_validation, labels=as.numeric(validation_data$y)-1)
performance(glm_pred_validation, 'auc')@y.values #GLM Validation AUC

stacking_prob_validation <- 1 - predict(stack_xgb, validation_data, type = "prob") #Gives predicted probability of y = 1
stacking_pred_validation <- prediction(stacking_prob_validation, labels=as.numeric(validation_data$y)-1)
performance(stacking_pred_validation, 'auc')@y.values #Stacking Validation AUC
mean(predict(stack_xgb, validation_data, type = "raw") == validation_data$y) #Stacking Validation Accuracy

#Predict on the Toy Data
stacking_prob_test <- 1 - predict(stack_xgb, test_data, type = "prob") #Gives predicted probability of y = 1
write.csv(stacking_prob_test, 'A0074604J.csv')
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
